---
title: "**Predictors of Weight in Romanian Infants: Weight-for-age z-scores (WAZ)**"
output:
  html_document: 
    keep_md: yes
    toc: yes
  pdf_document: default
  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

# *Adrian Rus*

*(September 2023)*

## 1. **Research Hypotheses/Questions/Objectives**

The objective of the present study was to explore the relationship between infants' ***weight*** and various ***demographics*** (i.e., infants' mother location, marital status, social economic status, and age; and infants being born on term, age, anemia status, rank into the family and dietary patterns).

Identifying the connection between such factors could help practitioners in developing public health policies and interventions in Romania.

## 2. **Data Exploration**

### 2.1. Setting Up the Environment: R Packages Installation

```{r warning=FALSE}

# the warnings will not appear in the knitted output

# install.packages("tidyverse", repos = "https://cloud.r-project.org/")
# install.packages("moments")
# install.packages(c("lmtest", "car", "carData", "ggplot2", "dplyr"))
library(tidyverse)
library(lmtest)
library(car)
library(carData)
library(ggplot2)
library(dplyr)

```

### 2.2. Setting the Working Directory

#### 2.2.1. Check the Current Working Directory

```{r}
getwd()
```

#### 2.2.2. Set A New Working Directory

You can change it using `setwd()`.

```{r}
setwd("C:\\Users\\adria\\Documents\\AR_Research_WAZ\\WAZ_Data & Scripts\\1_WAZ_EDA_R_2023-9-4")
```

Check the working directory

```{r}
getwd()
```

List Files in Directory To see a list of all files in the current working directory, you can use `list.files()`.

```{r}
list.files()
```

### 2.3. Import and Vizualize the Dataset

```{r}
library(readr)
WAZdataShort <- read_csv("WAZdataShort.csv")
View(WAZdataShort)
```

### 2.3.1. Basic inspection of dataset

##### Check id the dataset is tidy (see the tidyverse). A dataset is tidy when:

1\.
Each variable forms a column.

2\.
Each observation forms a row.

3\.
Each type of observational unit forms a table.

#### 2.3.1.1. Each row should correspond to a single observation.

##### Use `head()` or `View()` to visually inspect the data.

View the ***first*** 6 rows

```{r}
head(WAZdataShort)

```

View the ***last*** 6 rows

```{r}
tail(WAZdataShort)
```

#### 2.3.1.2. Each column should contain only one type of data.

View the structure of dataset

```{r}
str(WAZdataShort)
```

```{r}
summary(WAZdataShort)
```

#### 2.3.1.3. Tidy data should have a consistent structure, but it can contain missing values.

```{r}
sum(is.na(WAZdataShort)) # it shows the missing values in the dataset
```

#### 2.3.1.4. Tidy data should ideally not have duplicated rows unless the repetition is meaningful.

```{r}
sum(duplicated(WAZdataShort))
```

#### 2.3.1.5. Column names should be clear and descriptive but not too lengthy.

They shouldn't contain spaces or special characters.
Column names: Lists the names of all the columns in the dataset by using the `names(my_data)` or `colnames(my_data)`

```{r}
names(WAZdataShort)
```

#### 2.3.1.6. Data Types

```{r}
class(WAZdataShort) # Tells you whether the object is a data frame, matrix, vector, list, etc.
```

```{r}
sapply(WAZdataShort, class) # Returns the class (type) of each column in the data frame.
```

#### 2.3.1.7. Data Types Data Dimensionality

```{r}
ncol(WAZdataShort) # Returns the number of columns.
nrow(WAZdataShort) # Returns the number of rows.
dim(WAZdataShort) # Returns the number of rows and columns in a data frame.
```

```         
```

```{r}
# Select only numerical columns
numerical_columns <- WAZdataShort[, sapply(WAZdataShort, is.numeric)]
head(numerical_columns)
View(numerical_columns)
```

```{r}
# Select only categorical columns (factor or character)
categorical_columns <- WAZdataShort[, sapply(WAZdataShort, function(col) is.factor(col) || is.character(col))]
head(categorical_columns)
View(categorical_columns)
```

### 2.3.4. Data Content

```{r}
unique(WAZdataShort$mAge2c) # Shows unique values in a specific column.
```

```{r}
unique_values_all_columns <- lapply(WAZdataShort, unique) # Allows to identify the unique data in all columns
str(unique_values_all_columns)
```

```{r}
table(WAZdataShort$mAge2c) # Shows the frequency of each unique value in a specific column.
```

## 3. **Dependent Variable (DV)** and **Data Processing**

**Goals**:

1.  Conduct some preliminary analysis to understand the characteristics of the DV (its structure, missing values, and basic statistics).

2.  Handling missing values, normalization, and transformation, if necessary.

**DV**: is weight-for-age *z*-scores (WAZ)

### 3.1. Visual check of the DV

Histogram for DV

```{r}
library(ggplot2)
ggplot(WAZdataShort, aes(x = cWageZ)) + 
  geom_histogram(bins = 50, col= "white") +
  ggtitle("Histogram")
```

Histogram adding the normal distribution curve

```{r}
# Calculate mean and standard deviation of the data
data_mean <- mean(WAZdataShort$cWageZ, na.rm = TRUE)
data_sd <- sd(WAZdataShort$cWageZ, na.rm = TRUE)

```

```{r}
# Create the ggplot
ggplot(WAZdataShort, aes(x = cWageZ)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, col= "white") + 
  stat_function(fun = dnorm, args = list(mean = data_mean, sd = data_sd), color = "red") +
  ggtitle("Histogram with Normal Distribution Curve")
```

**Boxplot**: To identify outliers and understand the data's spread.

```{r}
ggplot(WAZdataShort, aes(y = cWageZ)) + geom_boxplot()
```

```{r}
ggplot(WAZdataShort, aes(y = cWageZ)) + 
  geom_boxplot(outlier.colour = "red")

```

**Density Plot**: To visualize the distribution.

```{r}
ggplot(WAZdataShort, aes(x = cWageZ)) + geom_density()
```

**Q-Q Plot**: To assess normality.

```{r}
ggplot(WAZdataShort, aes(sample = cWageZ)) + stat_qq()

```

**Pair Plot**: To visualize relationships between multiple numerical variables, if applicable.

```{r}
# Filter only numeric columns
WAZdataShort_numeric <- WAZdataShort[, sapply(WAZdataShort, is.numeric)]

# Create the pairs plot
pairs(WAZdataShort_numeric)

```

### 3.2. Compute central tendency, dispersion, skewness, kurtosis, and normality (Shapiro-Wilk P-value)

```{r}
library(tidyverse)
library(moments)
```

```{r}
summary_stats <- WAZdataShort %>%
  summarise(
    mean = mean(cWageZ, na.rm = TRUE),
    median = median(cWageZ, na.rm = TRUE),
    variance = var(cWageZ, na.rm = TRUE),
    sd = sd(cWageZ, na.rm = TRUE),
    skewness = skewness(cWageZ, na.rm = TRUE),
    kurtosis = kurtosis(cWageZ, na.rm = TRUE),
    shapiro_p_value = shapiro.test(cWageZ)$p.value
  )

# View the summary statistics
summary_stats
```

### 3.3. Other statistics used for identifying outliers

**Coefficient of Variation**: To compare the degree of variation if you have more than one DV.

```{r}
sd(WAZdataShort$cWageZ, na.rm = TRUE) / mean(WAZdataShort$cWageZ, na.rm = TRUE)
```

**Percentiles**: 25th, 50th, 75th, and other percentiles to understand distribution.

```{r}
quantile(WAZdataShort$cWageZ, probs = c(0.25, 0.5, 0.75), na.rm = TRUE)

```

**Confidence Intervals**: For mean or median.

```{r}
t.test(WAZdataShort$cWageZ)$conf.int  # For mean

```

**Tukey's method**: it uses the Interquartile Range (IQR) to identify outliers.

```{r}
# Load the dplyr package
library(dplyr)

# Tukey's Method to find outliers
Q1 <- quantile(WAZdataShort$cWageZ, 0.25)
Q3 <- quantile(WAZdataShort$cWageZ, 0.75)
IQR <- Q3 - Q1

outliers_Tukey <- WAZdataShort %>% 
  filter(cWageZ < (Q1 - 1.5 * IQR) | cWageZ > (Q3 + 1.5 * IQR))

# View the outliers
print(outliers_Tukey)

```

Group the outliers by **`cID`**

```{r}

library(dplyr)

# Grouping outliers by cID
grouped_outliers_Tukey <- outliers_Tukey %>%
  group_by(cID) %>%
  summarize(
    Num_Outliers = n(),  # Number of outliers per cID
    Min_cWageZ = min(cWageZ, na.rm = TRUE),  # Minimum value of cWageZ for each cID
    Max_cWageZ = max(cWageZ, na.rm = TRUE)  # Maximum value of cWageZ for each cID
  )

# View the grouped outliers
print(grouped_outliers_Tukey)

```

**Z-score Method**: used to find outliers

```{r}
# Load the dplyr package if not already loaded
# library(dplyr)

# Z-score Method to find outliers
mean_val <- mean(WAZdataShort$cWageZ, na.rm = TRUE)
sd_val <- sd(WAZdataShort$cWageZ, na.rm = TRUE)

outliers_Zscore <- WAZdataShort %>% 
  mutate(z_score = (cWageZ - mean_val) / sd_val) %>% 
  filter(abs(z_score) > 3)

# View the outliers
print(outliers_Zscore)

```

Group the outliers by **`cID`**

```{r}

#library(tidyr)

# Grouping outliers by cID using Z-score method and listing all outliers
grouped_outliers_Zscore_listed <- outliers_Zscore %>%
  group_by(cID) %>%
  summarize(
    Num_Outliers = n(),  # Number of outliers per cID
    Min_cWageZ = min(cWageZ, na.rm = TRUE),  # Minimum value of cWageZ for each cID
    Max_cWageZ = max(cWageZ, na.rm = TRUE),  # Maximum value of cWageZ for each cID
    All_Outliers = list(cWageZ)  # List all individual outlier values for each cID
  )

# View the grouped outliers
print(grouped_outliers_Zscore_listed)



```

**The Modified Z-score**: it uses the median and the Median Absolute Deviation (MAD) instead of the mean and standard deviation.
A common threshold is a Modified Z-score of +/- 3.5.

```{r}
# Load the dplyr package
# library(dplyr)

# Modified Z-score Method to find outliers
median_val <- median(WAZdataShort$cWageZ, na.rm = TRUE)
mad_val <- mad(WAZdataShort$cWageZ, na.rm = TRUE)

outliers_ModZscore <- WAZdataShort %>% 
  mutate(mod_z_score = 0.6745 * (cWageZ - median_val) / mad_val) %>% 
  filter(abs(mod_z_score) > 3.5)

# View the outliers
print(outliers_ModZscore)
```

### 3.4. Comparative Analysis of Outlier Detection Methods: Tukey's IQR, Z-Score, and Modified Z-Score

Among the methods presented for identifying outliers---Tukey's method, Z-score method, and Modified Z-score method---each has its own advantages and limitations, and the choice often depends on the specific context and requirements of the study.
However, Tukey's method using the interquartile range (IQR) is perhaps the most commonly used in exploratory data analysis for the following reasons:

#### **Advantages of Tukey's Method:**

1.  **Robust to Skewness**: Tukey's method doesn't assume that the data is normally distributed, making it suitable for skewed distributions.
2.  **Simple Interpretation**: The IQR-based rule is easy to understand and communicate, even for those without a strong statistical background.
3.  **Quick Computation**: It's computationally inexpensive and fast to calculate, which is useful for large datasets.
4.  **Incorporated into Boxplots**: Tukey's method is implicitly used in boxplots, a common visualization technique, making it convenient to both visualize and identify outliers simultaneously.
5.  **Less Sensitive to Extreme Value**s: Since it's based on the median and IQR, rather than the mean and standard deviation, it's less sensitive to extreme values.

#### When to Consider Other Methods:

1.  **Z-Score Method**: If you know or have reason to believe that your data is normally distributed, the Z-score method can be more appropriate. It's also more suited for studies where you are comparing the variability of different datasets.
2.  **Modified Z-Score Method**: If you want a compromise between the robustness of Tukey's method and the distributional assumptions of the Z-score method, the Modified Z-score using the median and the Median Absolute Deviation (MAD) can be a good choice.

In summary, the choice of method often depends on the specific characteristics of your data and what you know about its underlying distribution, as well as the questions you are trying to answer with your analysis.

### 3.5. Create a new version of DV without outliers.

#### Create a new DV: Tukey's method

```{r}

# Tukey's Method
Q1 <- quantile(WAZdataShort$cWageZ, 0.25, na.rm = TRUE)
Q3 <- quantile(WAZdataShort$cWageZ, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1

lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

# Add the new variable directly into the original data frame
WAZdataShort <- WAZdataShort %>%
  mutate(
    cWageZ_no_outliers = ifelse(cWageZ < lower_bound | cWageZ > upper_bound, NA, cWageZ)
  )

# View the data with the new variable
print(WAZdataShort)

```

#### Compare cWageZ to cWageZ_no_outliers (Tuker)

```{r}
# Load the ggplot2 package 
# library(ggplot2)

# Create a boxplot for the original DV (cWageZ)
ggplot(WAZdataShort, aes(y = cWageZ)) +
  geom_boxplot() +
  ggtitle("Boxplot of Original DV (cWageZ)")

# Create a boxplot for the new DV (cWageZ_no_outliers)
ggplot(WAZdataShort, aes(y = cWageZ_no_outliers)) +
  geom_boxplot() +
  ggtitle("Boxplot of New DV (cWageZ_no_outliers)")

```

```{r}
# Re-calculate Q1, Q3, and IQR for cWageZ_no_outliers
Q1_new <- quantile(WAZdataShort$cWageZ_no_outliers, 0.25, na.rm = TRUE)
Q3_new <- quantile(WAZdataShort$cWageZ_no_outliers, 0.75, na.rm = TRUE)
IQR_new <- Q3_new - Q1_new

# Define new lower and upper bounds
lower_bound_new <- Q1_new - 1.5 * IQR_new
upper_bound_new <- Q3_new + 1.5 * IQR_new

# Remove new outliers
WAZdataShort <- WAZdataShort %>%
  mutate(
    cWageZ_no_outliers = ifelse(cWageZ_no_outliers < lower_bound_new | cWageZ_no_outliers > upper_bound_new, NA, cWageZ_no_outliers)
  )

# Create a boxplot for the new DV (cWageZ_no_outliers)
ggplot(WAZdataShort, aes(y = cWageZ_no_outliers)) +
  geom_boxplot() +
  ggtitle("Boxplot of New DV without Outliers (cWageZ_no_outliers)")
```

#### Create a new DV: Z-score method

```{r}
library(dplyr)
library(rlang)

# Function to iteratively remove outliers based on Z-score
iterative_zscore_removal <- function(df, col_name) {
  col_sym <- sym(col_name)
  
  df <- df %>% mutate(Outlier = FALSE)
  
  repeat {
    mean_val <- mean(df %>% pull(!!col_sym), na.rm = TRUE)
    sd_val <- sd(df %>% pull(!!col_sym), na.rm = TRUE)
    
    # Identify outliers
    df <- df %>% mutate(Outlier = abs((!!col_sym - mean_val) / sd_val) > 3)
    
    # Check if any outliers exist
    if (sum(df$Outlier, na.rm = TRUE) == 0) {
      break
    }
    
    # Replace outliers with NA
    df <- df %>% mutate(
      !!col_name := if_else(Outlier, NA_real_, !!col_sym)
    )
  }
  
  df$Outlier <- NULL  # Remove the Outlier column
  return(df)
}

# Apply the function to remove outliers iteratively from cWageZ
WAZdataShort <- iterative_zscore_removal(WAZdataShort, "cWageZ")

# Create a new column with non-outlier values
WAZdataShort <- WAZdataShort %>%
  mutate(
    cWageZ_no_outliers_robust = cWageZ
  )

# View the updated data
View(WAZdataShort)

```

#### Compare cWageZ to cWageZ_no_outliers (Tuker) and cWageZ_no_outliers_robust (Z-score)

```{r}
# Load required libraries
# library(ggplot2)
# library(tidyr)

# Reshape the data into long format
WAZdataLong <- WAZdataShort %>%
  select(cWageZ, cWageZ_no_outliers, cWageZ_no_outliers_robust) %>%
  gather(key = "Variable", value = "Value")

# Create the boxplot
ggplot(WAZdataLong, aes(x = Variable, y = Value)) +
  geom_boxplot() +
  ggtitle("Comparison of Original and Cleaned DVs") +
  xlab("Variable") +
  ylab("Value")

```

### 3.6. Save the dataset to contain the new version of IV & DV with and without outliers.

```{r}

# save the data
# write.csv(WAZdataShort, "WAZdata_short_no_outliers.csv")

```

### 3.7. Importing the data used for running the regression models

```{r}

library(tidyverse)

# Read the data
WAZdata_short_no_outliers <- read_csv("WAZdata_short_no_outliers.csv")

# Create the tibble (this step might be redundant if WAZdata_short_no_outliers is already a tibble)

WAZ_d_modeling <- tibble(WAZdata_short_no_outliers)

print(WAZ_d_modeling)
View(WAZ_d_modeling)
str(WAZ_d_modeling)

```

## 4. **Independent Variables (IV)** and **Data Processing**

**Goals**:

1.  Conduct some preliminary analysis to understand the characteristics of the IV (its structure, missing values, and basic statistics).

2.  Handling missing values, normalization, and transformation, if necessary.

**IV: Model 1** - selected based on the previous published paper --\> Location, cBirthT, cAgeM3c, mMarital2, Anemia2, mSES3c, cMDD, cRank3c, mAge

**IV: Model 2** - selectd based on correlation with DV --\> Model 1 + cHageZ

### 4.1. Visual check of the IV (for numerical Variables)

```{r}

# Load the ggplot2 package 
library(ggplot2)


# Create a boxplot
ggplot(WAZ_d_modeling, aes(y = mAge)) +
  geom_boxplot() +
  ggtitle("Boxplot of (Mothers' Age; mAge)")

```

```{r}

# Load the ggplot2 package 
library(ggplot2)

# Create a boxplot
ggplot(WAZ_d_modeling, aes(y = cHageZ)) +
  geom_boxplot() +
  ggtitle("Boxplot of (cHageZ)")

```

### 4.2. Compute central tendency, dispersion, skewness, kurtosis, and normality (Shapiro-Wilk P-value)

```{r}

# Load the necessary libraries
library(moments)
library(dplyr)

# Calculate summary statistics
summary_stats <- WAZ_d_modeling %>%
  summarise(
    mean_mAge = mean(mAge, na.rm = TRUE),
    median_mAge = median(mAge, na.rm = TRUE),
    variance_mAge = var(mAge, na.rm = TRUE),
    sd_mAge = sd(mAge, na.rm = TRUE),
    skewness_mAge = skewness(mAge, na.rm = TRUE),
    kurtosis_mAge = kurtosis(mAge, na.rm = TRUE),
    shapiro_p_value_mAge = shapiro.test(mAge)$p.value,
    
    mean_cHageZ = mean(cHageZ, na.rm = TRUE),
    median_cHageZ = median(cHageZ, na.rm = TRUE),
    variance_cHageZ = var(cHageZ, na.rm = TRUE),
    sd_cHageZ = sd(cHageZ, na.rm = TRUE),
    skewness_cHageZ = skewness(cHageZ, na.rm = TRUE),
    kurtosis_cHageZ = kurtosis(cHageZ, na.rm = TRUE),
    shapiro_p_value_cHageZ = shapiro.test(cHageZ)$p.value
  )

# View the summary statistics
summary_stats

```

```{r}

# Load the necessary libraries
# install.packages("moments")
library(moments)
library(dplyr)
library(tidyr)

# Calculate summary statistics
summary_stats <- WAZ_d_modeling %>%
  summarise(
    mean_mAge = mean(mAge, na.rm = TRUE),
    median_mAge = median(mAge, na.rm = TRUE),
    variance_mAge = var(mAge, na.rm = TRUE),
    sd_mAge = sd(mAge, na.rm = TRUE),
    skewness_mAge = skewness(mAge, na.rm = TRUE),
    kurtosis_mAge = kurtosis(mAge, na.rm = TRUE),
    shapiro_p_value_mAge = shapiro.test(mAge)$p.value,
    
    mean_cHageZ = mean(cHageZ, na.rm = TRUE),
    median_cHageZ = median(cHageZ, na.rm = TRUE),
    variance_cHageZ = var(cHageZ, na.rm = TRUE),
    sd_cHageZ = sd(cHageZ, na.rm = TRUE),
    skewness_cHageZ = skewness(cHageZ, na.rm = TRUE),
    kurtosis_cHageZ = kurtosis(cHageZ, na.rm = TRUE),
    shapiro_p_value_cHageZ = shapiro.test(cHageZ)$p.value
  ) %>%
  gather(key = "Statistic", value = "Value")

# View the summary statistics
summary_stats

```

```{r}

# Load the necessary libraries
# install.packages("moments")
library(moments)
library(dplyr)
library(tidyr)

# Calculate summary statistics
summary_stats <- WAZ_d_modeling %>%
  summarise(
    mean_mAge = round(mean(mAge, na.rm = TRUE), 4),
    median_mAge = round(median(mAge, na.rm = TRUE), 4),
    variance_mAge = round(var(mAge, na.rm = TRUE), 4),
    sd_mAge = round(sd(mAge, na.rm = TRUE), 4),
    skewness_mAge = round(skewness(mAge, na.rm = TRUE), 4),
    kurtosis_mAge = round(kurtosis(mAge, na.rm = TRUE), 4),
    shapiro_p_value_mAge = round(shapiro.test(mAge)$p.value, 4),
    
    mean_cHageZ = round(mean(cHageZ, na.rm = TRUE), 4),
    median_cHageZ = round(median(cHageZ, na.rm = TRUE), 4),
    variance_cHageZ = round(var(cHageZ, na.rm = TRUE), 4),
    sd_cHageZ = round(sd(cHageZ, na.rm = TRUE), 4),
    skewness_cHageZ = round(skewness(cHageZ, na.rm = TRUE), 4),
    kurtosis_cHageZ = round(kurtosis(cHageZ, na.rm = TRUE), 4),
    shapiro_p_value_cHageZ = round(shapiro.test(cHageZ)$p.value, 4)
  ) %>%
  gather(key = "Statistic", value = "Value")

# View the summary statistics
summary_stats


```

### 4.3. Create new variables by excluding the outliers using

**Tukey's method**: it uses the Interquartile Range (IQR) to identify outliers.

```{r}

# Tukey's Method for mAge
Q1_mAge <- quantile(WAZ_d_modeling$mAge, 0.25, na.rm = TRUE)
Q3_mAge <- quantile(WAZ_d_modeling$mAge, 0.75, na.rm = TRUE)
IQR_mAge <- Q3_mAge - Q1_mAge

lower_bound_mAge <- Q1_mAge - 1.5 * IQR_mAge
upper_bound_mAge <- Q3_mAge + 1.5 * IQR_mAge

# Tukey's Method for cHageZ
Q1_cHageZ <- quantile(WAZ_d_modeling$cHageZ, 0.25, na.rm = TRUE)
Q3_cHageZ <- quantile(WAZ_d_modeling$cHageZ, 0.75, na.rm = TRUE)
IQR_cHageZ <- Q3_cHageZ - Q1_cHageZ

lower_bound_cHageZ <- Q1_cHageZ - 1.5 * IQR_cHageZ
upper_bound_cHageZ <- Q3_cHageZ + 1.5 * IQR_cHageZ

# Add the new variables directly into the original data frame
WAZ_d_modeling <- WAZ_d_modeling %>%
  mutate(
    mAge_no_outliers = ifelse(mAge < lower_bound_mAge | mAge > upper_bound_mAge, NA, mAge),
    cHageZ_no_outliers = ifelse(cHageZ < lower_bound_cHageZ | cHageZ > upper_bound_cHageZ, NA, cHageZ)
  )

# View the data with the new variables
print(WAZ_d_modeling)
View(WAZ_d_modeling)
str(WAZ_d_modeling)

```

### 4.4. Visual check of the new IV

**mAge_no_outliers**

```{r}

# Load the ggplot2 package 
library(ggplot2)

# Create a boxplot
ggplot(WAZ_d_modeling, aes(y = mAge_no_outliers)) +
  geom_boxplot() +
  ggtitle("Boxplot of (Mothers' Age; mAge_no_outliers)")

```

**cHageZ_no_outliers**

```{r}

# Load the ggplot2 package 
library(ggplot2)

# Create a boxplot
ggplot(WAZ_d_modeling, aes(y = cHageZ_no_outliers)) +
  geom_boxplot() +
  ggtitle("Boxplot of (cHageZ_no_outliers)")

```

### 4.5. Continue the process of eliminating the outliers from cHageZ_no_outliers using

**Tukey's method**: it uses the Interquartile Range (IQR) to identify outliers.

```{r}

# Initialize a flag for the while loop
has_outliers <- TRUE

while(has_outliers) {
  # Tukey's Method for cHageZ_no_outliers
  Q1 <- quantile(WAZ_d_modeling$cHageZ_no_outliers, 0.25, na.rm = TRUE)
  Q3 <- quantile(WAZ_d_modeling$cHageZ_no_outliers, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  
  # Detect outliers
  outliers <- WAZ_d_modeling %>% 
    filter((cHageZ_no_outliers < lower_bound | cHageZ_no_outliers > upper_bound), !is.na(cHageZ_no_outliers))
  
  # Check if any outliers are detected
  if (nrow(outliers) > 0) {
    # Remove outliers
    WAZ_d_modeling <- WAZ_d_modeling %>%
      mutate(
        cHageZ_no_outliers = ifelse(cHageZ_no_outliers < lower_bound | cHageZ_no_outliers > upper_bound, NA, cHageZ_no_outliers)
      )
  } else {
    # Exit the loop if no outliers
    has_outliers <- FALSE
  }
}


```

**cHageZ_no_outliers**

```{r}

# Load the ggplot2 package 
library(ggplot2)

# Create a boxplot
ggplot(WAZ_d_modeling, aes(y = cHageZ_no_outliers)) +
  geom_boxplot() +
  ggtitle("Boxplot of (cHageZ_no_outliers)")

```

### 4.6. Save the dataset to contain the new version of DV and IV without outliers.

```{r}

# save the data
# write.csv(WAZ_d_modeling, "WAZ_d_modeling_final.csv", row.names = FALSE)

```

## 5. **Statistical Testing**

**Goal**: Employ appropriate statistical tests or models to analyze the relationship between the IV and DV.

### 5.1. Importing the data used for running the regression models

```{r}

# Read the data
library(readr)
WAZ_d_modeling_final <- read_csv("WAZ_d_modeling_final.csv")
print(WAZ_d_modeling_final)
View(WAZ_d_modeling_final)
str(WAZ_d_modeling_final)

```

### 5.2. Model 0: The intercept-only model (null model)

```{r}

library(tidyverse)

# Run the intercept-only model (null model)
Model_0 <- lm(cWageZ ~ 1, data = WAZ_d_modeling_final)

# View the summary of the model
summary(Model_0)

```

### 5.3. Set the levels of IV

```{r}

# Explicitly convert all variables to factors
WAZ_d_modeling_final$Location <- as.factor(WAZ_d_modeling_final$Location)
WAZ_d_modeling_final$cBirthT <- as.factor(WAZ_d_modeling_final$cBirthT)
WAZ_d_modeling_final$cAgeM3c <- as.factor(WAZ_d_modeling_final$cAgeM3c)
WAZ_d_modeling_final$mMarital2 <- as.factor(WAZ_d_modeling_final$mMarital2)
WAZ_d_modeling_final$Anemia2 <- as.factor(WAZ_d_modeling_final$Anemia2)
WAZ_d_modeling_final$mSES3c <- as.factor(WAZ_d_modeling_final$mSES3c)
WAZ_d_modeling_final$cMDD <- as.factor(WAZ_d_modeling_final$cMDD)
WAZ_d_modeling_final$cRank3c <- as.factor(WAZ_d_modeling_final$cRank3c)

# Now proceed to relevel the variables
WAZ_d_modeling_final$Location <- relevel(WAZ_d_modeling_final$Location, ref = "Urban")
WAZ_d_modeling_final$cBirthT <- relevel(WAZ_d_modeling_final$cBirthT, ref = "At term (37-40 weeks)")
WAZ_d_modeling_final$cAgeM3c <- relevel(WAZ_d_modeling_final$cAgeM3c, ref = "6-11 months")
WAZ_d_modeling_final$mMarital2 <- relevel(WAZ_d_modeling_final$mMarital2, ref = "Married")
WAZ_d_modeling_final$Anemia2 <- relevel(WAZ_d_modeling_final$Anemia2, ref = "No (11> g/dl)")
WAZ_d_modeling_final$mSES3c <- relevel(WAZ_d_modeling_final$mSES3c, ref = "Low")
WAZ_d_modeling_final$cMDD <- relevel(WAZ_d_modeling_final$cMDD, ref = "Yes")
WAZ_d_modeling_final$cRank3c <- relevel(WAZ_d_modeling_final$cRank3c, ref = "First")


```

### 5.4. Model 1:

**The structure of the model**

**DV**: cWageZ

**IV**: Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge

```{r}
# Model 1
Model_1 <- lm(cWageZ ~ Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge, data = WAZ_d_modeling_final)

# View the summary of the model
summary(Model_1)

# Perform Omnibus Test
anova(Model_1)

# Get Confidence Intervals
confint(Model_1)

```

**Model 1**: Load packages

```{r}

library(lmtest)
library(car)
library(carData)
library(ggplot2)
library(dplyr)

```

**Model 1**: Diagnostic -- Plotting Linearity and Normality

```{r}

Model_1 <- lm(cWageZ ~ Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge, data = WAZ_d_modeling_final)

# Added Variable Plots for assessing linearity
avPlots(Model_1)
 
# Q-Q Plot for assessing normality of residuals
qqPlot(Model_1, id.n = 0)

```

**Model 1**: Diagnostic Tests for Model Assumptions: Homoscedasticity, Independence, and Multicollinearity

```{r}

# Load necessary libraries
library(car) # for vif(), bptest(), and dwtest()
library(lmtest) # for dwtest()

# Fit the model
Model_1 <- lm(cWageZ ~ Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge, data = WAZ_d_modeling_final)

# Run Breusch-Pagan test and capture p-value
bp_p_value <- bptest(Model_1)$p.value

# Run Durbin-Watson test and capture p-value
dw_p_value <- dwtest(Model_1)$p.value

# Calculate VIF for all predictors
individual_vif <- vif(Model_1)

# Calculate average VIF for all predictors
avg_vif <- mean(individual_vif)

# Round all values to 4 decimal places
bp_p_value <- round(bp_p_value, 4)
dw_p_value <- round(dw_p_value, 4)
avg_vif <- round(avg_vif, 4)
individual_vif <- round(individual_vif, 4)

# Create a table to display the diagnostics
diagnostics_table <- data.frame(
  Category = rep("General", 3),
  Test = c("Breusch-Pagan p-value", "Durbin-Watson p-value", "Average VIF"),
  Value = c(bp_p_value, dw_p_value, avg_vif)
)

# Create a table to display the general diagnostics
diagnostics_table <- data.frame(
  Test = c("Breusch-Pagan p-value", "Durbin-Watson p-value", "Average VIF"),
  Value = c(bp_p_value, dw_p_value, avg_vif)
)

# Display the diagnostics table
print("General Diagnostics:")
print(diagnostics_table)

# Create a table to display individual VIF values
vif_table <- data.frame(
  Test = paste("VIF for", names(individual_vif)),
  Value = individual_vif
)

# Display the VIF table
print("Individual VIFs:")
print(vif_table)


```

### 5.5 Model 1b: Variables with no outliers were included

**The structure of the model**

**DV**: cWageZ_no_outliers

**IV**: Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge_no_outliers

```{r}
# Model 1b
Model_1b_no_outliers <- lm(cWageZ_no_outliers ~ Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge_no_outliers, data = WAZ_d_modeling_final)

# View the summary of the model
summary(Model_1b_no_outliers)

# Perform Omnibus Test
anova(Model_1b_no_outliers)

# Get Confidence Intervals
confint(Model_1b_no_outliers)

```

### 5.6. Model 2:

**The sructure of the model**

**DV**: cWageZ

**IV**: Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge + cHageZ

```{r}

# Model 2
Model_2 <- lm(cWageZ ~ Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge + cHageZ, data = WAZ_d_modeling_final)

# View the summary of the model
summary(Model_2)

# Perform Omnibus Test
anova(Model_2)

# Get Confidence Intervals
confint(Model_2)

```

**Model 2**: Diagnostic -- Plotting Linearity and Normality

```{r}

# Model 2
Model_2 <- lm(cWageZ ~ Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge + cHageZ, data = WAZ_d_modeling_final)

# Added Variable Plots for assessing linearity
avPlots(Model_2)
 
# Q-Q Plot for assessing normality of residuals
qqPlot(Model_2, id.n = 0)

```

**Model 2**: Diagnostic Tests for Model Assumptions: Homoscedasticity, Independence, and Multicollinearity

```{r}
# Load necessary libraries
library(car)
library(lmtest)

# Define the model
Model_2 <- lm(cWageZ ~ Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge + cHageZ, data = WAZ_d_modeling_final)

# Breusch-Pagan Test for assessing homoscedasticity
bp_test <- bptest(Model_2)

# Durbin-Watson Test for assessing independence of residuals
dw_test <- dwtest(Model_2)

# Variance Inflation Factors (VIF) for assessing multicollinearity
vif_values <- vif(Model_2)

# Create a table to store the results
diagnostic_table <- tibble(
  Test = c("Breusch-Pagan Test p-value", "Durbin-Watson Test p-value"),
  Value = c(bp_test$p.value, dw_test$p.value),
  Notes = c("Assessing homoscedasticity", "Assessing independence of residuals")
)

# Add VIF to the table
vif_table <- as_tibble(vif_values, rownames = "Variable")
colnames(vif_table) <- c("Test", "Value")
vif_table$Notes <- "Assessing multicollinearity"

# Combine tables
final_table <- bind_rows(diagnostic_table, vif_table)

# Print the final table
print(final_table)

```

### 5.7. Model 2b: Variables with no outliers were inclued

**The sructure of the model**

**DV**: cWageZ

**IV**: Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge + cHageZ

```{r}

# Model 2b
Model_2b_no_outliers <- lm(cWageZ_no_outliers ~ Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge_no_outliers + cHageZ_no_outliers, data = WAZ_d_modeling_final)

# View the summary of the model
summary(Model_2b_no_outliers)

# Perform Omnibus Test
anova(Model_2b_no_outliers)

# Get Confidence Intervals
confint(Model_2b_no_outliers)

```

### 5.8 Compare Model 1 and Model 2

Models comparison

```{r}

# Summary statistics for Model_1
summary_1 <- summary(Model_1)
print(paste("AIC for Model 1: ", round(AIC(Model_1), 4)))
print(paste("BIC for Model 1: ", round(BIC(Model_1), 4)))
print(paste("Adjusted R-squared for Model 1: ", round(summary_1$adj.r.squared, 4)))

# Extract p-values and round them to 4 digits
p_values_1 <- round(summary_1$coefficients[, "Pr(>|t|)"], 4)

# Summary statistics for Model_2
summary_2 <- summary(Model_2)
print(paste("AIC for Model 2: ", round(AIC(Model_2), 4)))
print(paste("BIC for Model 2: ", round(BIC(Model_2), 4)))
print(paste("Adjusted R-squared for Model 2: ", round(summary_2$adj.r.squared, 4)))

# Extract p-values and round them to 4 digits
p_values_2 <- round(summary_2$coefficients[, "Pr(>|t|)"], 4)

# ANOVA test between Model 1 and Model 2
anova_result <- anova(Model_1, Model_2)
anova_p_value <- round(anova_result$"Pr(>F)"[2], 4)
print(paste("ANOVA p-value comparison between Model 1 and Model 2: ", anova_p_value))

```

Create a table with the output

```{r}

# Summary statistics for Model_1
summary_1 <- summary(Model_1)
aic_1 <- round(AIC(Model_1), 4)
bic_1 <- round(BIC(Model_1), 4)
adj_r2_1 <- round(summary_1$adj.r.squared, 4)

# Summary statistics for Model_2
summary_2 <- summary(Model_2)
aic_2 <- round(AIC(Model_2), 4)
bic_2 <- round(BIC(Model_2), 4)
adj_r2_2 <- round(summary_2$adj.r.squared, 4)

# ANOVA test between Model 1 and Model 2
anova_result <- anova(Model_1, Model_2)
anova_p_value <- round(anova_result$"Pr(>F)"[2], 4)

# Create a data frame to hold the metrics
model_comparison <- data.frame(
  Metric = c("AIC", "BIC", "Adjusted R-squared", "ANOVA p-value"),
  Model_1 = c(aic_1, bic_1, adj_r2_1, "N/A"),
  Model_2 = c(aic_2, bic_2, adj_r2_2, anova_p_value)
)

# Display the table
print(model_comparison)


```

### 5.9. Model 3: Only statistical significant variables from Model 2 included

**The structure of the model**

**DV**: cWageZ

**IV**: cBirthT + cAgeM3c + Anemia2 + mSES3c + cHageZ

```{r}

# Model 3
Model_3 <- lm(cWageZ ~ cBirthT + cAgeM3c + Anemia2 + mSES3c + cHageZ, data = WAZ_d_modeling_final)

# View the summary of the model
summary(Model_3)

# Perform Omnibus Test
anova(Model_3)

# Get Confidence Intervals
confint(Model_3)

```

### 5.10. Model 3b: Only statistical significant variables with no outliers from Model 2 included

**The structure of the model**

**DV**: cWageZ_no_outliers

**IV**:cBirthT + cAgeM3c + Anemia2 + mSES3c + cHageZ_no_outliers

```{r}
# Model 3b
Model_3b_no_outliers <- lm(cWageZ_no_outliers ~ cBirthT + cAgeM3c + Anemia2 + mSES3c + cHageZ_no_outliers, data = WAZ_d_modeling_final)

# View the summary of the model
summary(Model_3b_no_outliers)

# Perform Omnibus Test
anova(Model_3b_no_outliers)

# Get Confidence Intervals
confint(Model_3b_no_outliers)

```

### 5.11. Comprehensive Regression Models for Analyzing Wage Determinants: Including and Excluding Outliers

```{r}

# Model 1
Model_1 <- lm(cWageZ ~ Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge, data = WAZ_d_modeling_final)

# Model 1b without outliers
Model_1b_no_outliers <- lm(cWageZ_no_outliers ~ Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge_no_outliers, data = WAZ_d_modeling_final)

# Model 2
Model_2 <- lm(cWageZ ~ Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge + cHageZ, data = WAZ_d_modeling_final)

# Model 2b without outliers
Model_2b_no_outliers <- lm(cWageZ_no_outliers ~ Location + cBirthT + cAgeM3c + mMarital2 + Anemia2 + mSES3c + cMDD + cRank3c + mAge_no_outliers + cHageZ_no_outliers, data = WAZ_d_modeling_final)

# Model 3
Model_3 <- lm(cWageZ ~ cBirthT + cAgeM3c + Anemia2 + mSES3c + cHageZ, data = WAZ_d_modeling_final)

# Model 3b without outliers
Model_3b_no_outliers <- lm(cWageZ_no_outliers ~ cBirthT + cAgeM3c + Anemia2 + mSES3c + cHageZ_no_outliers, data = WAZ_d_modeling_final)



# Calculate adjusted R-squared, AIC, and BIC
summary_stats <- data.frame(
  Model = c("Model_1", "Model_1b_no_outliers", "Model_2", "Model_2b_no_outliers", "Model_3", "Model_3b_no_outliers"),
  Adj_R2 = c(
    summary(Model_1)$adj.r.squared,
    summary(Model_1b_no_outliers)$adj.r.squared,
    summary(Model_2)$adj.r.squared,
    summary(Model_2b_no_outliers)$adj.r.squared,
    summary(Model_3)$adj.r.squared,
    summary(Model_3b_no_outliers)$adj.r.squared
  ),
  AIC = c(
    AIC(Model_1),
    AIC(Model_1b_no_outliers),
    AIC(Model_2),
    AIC(Model_2b_no_outliers),
    AIC(Model_3),
    AIC(Model_3b_no_outliers)
  ),
  BIC = c(
    BIC(Model_1),
    BIC(Model_1b_no_outliers),
    BIC(Model_2),
    BIC(Model_2b_no_outliers),
    BIC(Model_3),
    BIC(Model_3b_no_outliers)
  )
)

# ANOVA test (only for nested models)
anova_test_1_2 <- anova(Model_1, Model_2)
anova_test_1_3 <- anova(Model_1, Model_3)

# Display the tables
print("Summary Statistics")
print(summary_stats)
print("ANOVA for Model 1 and Model 2")
print(anova_test_1_2)
print("ANOVA for Model 1 and Model 3")
print(anova_test_1_3)


```

## 6. **Reporting and Interpreting the Outcomes**

The report of the outcomes and interpretation will be completed in a different document.
